Analysis of Cartesian Genetic Programming's Evolutionary Mechanisms

### 2013-11-12 CGP Journal/main.tex
Fixed "footprint"->"sematic" in the sudocode
First hack of reviewer suggestions

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering.

@@

As both output locations and information flow in the DAG are evolvable, often
only a tiny fraction of the genome participates in creating the
output values.  These nodes are referred to as ``active,'' with the nodes not
being used to create output values referred to as ``inactive.''  Inactive nodes
allow CGP individuals to drift genetically, as they can be mutated without affecting
the fitness of the individual.  This genetic drift can then be incorporated, as
mutation can change the DAG structure causing previously inactive nodes to become
active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.  Mutation
is commonly performed by mutating each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.  Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

@@

The number of wasted evaluations is
directly related to the mutation rate and the number of active genes.  With
sufficiently high mutation rates and enough active genes, the probability of
creating an offspring which is actively identical to its parent is very low.
Yet such a configuration may make it difficult for CGP to optimize, as mutation
is likely to change many active genes in each application.  As was discussed
in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
mutation rate.

\subsection{Skip}
The \emph{Skip} method for avoiding duplicate evaluations involves the least
amount of modification from CGP's normal behavior.  After an offspring is produced
and its set of active nodes is determined but before it is evaluated, it is compared
with its parent.  Each gene in each active node is compared for equivalence with
the corresponding gene in the offspring's parent.  If all active genes are found to be equal
than the offspring is not evaluated and is instead given the same fitness
as its parent.
Note that since inactive genes are not compared for equivalence it is still possible
for offspring to genetically differ from their parents.

Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
it will produce identical individuals, with a potential reduction in evaluations.
If the mutation rate is high enough relative
to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
the probability of an offspring being actively identical to its parent is effectively zero.
When the mutation rate is low enough relative to the number of active genes, some number of
offspring will be actively identical to their parents, resulting in a reduction in evaluations
but no change in the evolutionary trajectory.

We would expect both intuitively and from our initial
experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
to the mutation rate than normal CGP.  We would also expect \emph{Skip} to use a lower
mutation rate than normal CGP when both are optimized to reduce evaluations.
The former comes from the fact that in normal CGP any mutation rate that has a significant
probability of creating offspring actively identical to their parents wastes
evaluations.  In \emph{Skip} there are only two penalties for reducing the
mutation rate.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
rates may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
exceptionally low mutation rates may spend a prohibitive amount of time
attempting to produce an evaluable offspring.
Note that these penalties also exist in normal CGP.

@@

Previous experimentation has shown that \emph{Accumulate} acts very similarly to
\emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
favors lower mutation rates when tuned to reduce evaluations to success. While at first these algorithms may appear
quite different, further consideration shows how they are similar.  In \emph{Skip}
any actively identical offspring that is produced is likely to be selected as we expect mutations to
active genes to more often reduce fitness than improve it.  In the next generation
this actively identical offspring is then mutated again, with its lineage likely
continuing until it finally does mutate one or more active genes.  At this point, if the
mutant is better it replaces the parent, otherwise the parent, which has been accumulating
mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
except \emph{Skip} acts over multiple generations instead of compressed into a single generation.

@@

This modification gives \emph{Single} three properties distinct from the other
forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
an active gene which differs from their parent.  This limits drift as the
mutated active gene must either be to an intron or represent another way to
code a solution of equal quality.  Second, as a benefit of forced changes,
\emph{Single} avoids the overhead of repeatedly generating individuals without
creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
rate parameter, effectively setting the mutation rate to $\frac{1}{a}$ for active
genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.

@@

\section{Genome Ordering}
\label{sec:ordering}
As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
to prevent cycles, by ensuring nodes only receive input from sources that
precede them in the genome.  This restriction does not limit CGP's ability
to represent DAGs, as all DAGs can be serialized to fit this requirement.
However, this representation likely has an impact on CGP's ability to evolve specific
DAGs~\cite{goldman:2013:ordering}.

@@

To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
created \emph{Reorder} and \emph{DAG} as methods for genome ordering, and defined
\emph{Normal} as CGP's historical ordering method.

@@

The first step in performing \emph{Reorder} on a genome is to create convenience
data structures to store direct connection relationships.  These structures are used
to determine in constant time all connections incident on a node, and 
can be built in order $O(AN)$ time, where $A$ is the arity of the nodes
and $N$ is the number of nodes.

With this preprocessing complete, the \Call{RandomSerialization}{} algorithm given in Figure~\ref{fig:reorder}
is used to assign each node a new location in the modified genome.
It starts by constructing the \emph{addable} set, which contains all nodes who's direct
dependencies have already been added.  Initially this contains only the input locations.
Once a node is assigned a location, all of the nodes that depend on its output
have that dependency marked as satisfied.  Once all of a nodes dependencies
are satisfied, it can be put into \emph{addable}.  In this way, nodes are
randomly removed from \emph{addable}, assigned the next possible location
in the genome, with new nodes added to \emph{addable} as they becomes semantically
viable.  Iteration ends when all nodes have been assigned new locations.
\Call{RandomSerialization}{} requires $O(AN)$, as there can be only $N$ iterations,
with each iteration updating a distinct subset of the $AN$ dependencies.

@@

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.  While this increase
makes finding a specific offspring that improves upon the parent less likely, the
general likelihood of producing an offspring that improves upon the parent may 
more than offset this drawback.

@@

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

\Call{IsDependent}{} is a non-repeated recursive depth
first search of the individual's DAG.  Search terminates as soon as a dependent
node is found, and information from previous searches is memoized to prevent
repeated search.  Each recursion level results in a node being added to $known$.
As such initial calls will likely result in lots of recursion, but subsequent
calls will find answers more quickly.  In the worst case, this algorithm may need
to examine all nodes in the genome to determine if $working$ is dependent on the
mutating node, giving \Call{IsDependent}{} a complexity of $O(AN)$.  In the worst case, this algorithm
may be called on all possible nodes in the genome to check their dependence.
Yet cumulatively those calls can only take $O(AN)$ time due to the non-repeating
nature of the search.  As a result, the expected worst case run time for \emph{DAG}
mutation is $O(mA^2N^2)$, where $m$ is the mutation rate.  We expect $mAN$
connection genes to be mutated, each requiring a worst case of $N$ dependency checks
each taking $A$ time.
This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
mutation, as individual mutations can be performed in constant time.

@@

\begin{table}
	\centering
	\begin{tabular}{|c|p{6cm}|}
	\hline
  \multicolumn{2}{|c|}{\textbf{Duplication}} \\ \hline
	\textbf{\emph{Skip}} & Whenever an actively identical offspring is detected, assign it its parent's fitness. \\ & \\
	\textbf{\emph{Accumulate}} & Repeatedly perform full genome mutation until \textbf{one or more} active genes change. \\ & \\
	\textbf{\emph{Single}} & Select and mutate individual genes until \textbf{exactly one} active gene changes. \\
	\hline \multicolumn{2}{|c|}{\textbf{Ordering}} \\ \hline
	\textbf{\emph{Normal}} & Nodes must take input from other nodes or inputs that precede them in the genome. \\ & \\
	\textbf{\emph{Reorder}} & Each generation the parent's node ordering is shuffled without effecting node behavior. \\ & \\
	\textbf{\emph{DAG}} & Connection genes can mutate to make any connection as long as no cycle is formed. \\
	  \hline
	\end{tabular}
	\caption{Short descriptions of each of the algorithm variants.}
	\label{tab:glossary}
\end{table}

@@

We focus on two parameter values in our configuration: mutation rate and genome size.
While there are other potential configurable parameters (population size, offspring size, etc),
we feel the CGP literature has converged on settings for those parameters (1, 4).
Furthermore, as duplication detection explicitly deals with the relationship between
mutation rate and number of active genes, and the number of active genes depends
on the genome size and ordering method, mutation rate and genome size seem to be
the most likely to impact results.

To set parameter values, we started by defining a grid of parameter values:
50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation rate.  Note that here a genome
size of 50 means there are 50 nodes, each composed of multiple genes, plus the
required genes to specify output locations.  Also note that a mutation rate of 0.05
is used here to mean each gene has a 5\% probability of being mutated in a single
application of mutation.  These ranges were chosen to completely cover the
range of previously used parameter settings.
The grid is formed by trying all possible mutation rates
with all possible genome sizes, resulting in 64 potential configurations for each
of the algorithm combinations.  Finally note that as \emph{Single} does not
use a mutation rate, only the eight configurations with different genome sizes were
used when \emph{Single} was employed.

To choose which parameter configuration to use with each algorithm combination
for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
and repeating.  For each algorithm combination and problem, we first ran all
potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
The parameter configurations are then sorted based on their median.  The best half are then
each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
before any were removed.  This iterative reduction allows us to focus tuning on
those parameter configurations most likely to be effective, with more information
gathered before choosing between high quality configurations.  In total this means
15,696 complete runs are required to set the parameter configuration for all
algorithm configurations for all problems.

@@

\begin{table*}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Rate} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline\hline

@@

As discussed in Section~\ref{sec:parameter}, the genome size and mutation rate
for each algorithm combination was extensively optimized.  In line with previous
research~\cite{miller:2006:redundancy}, this led to massive genome sizes for
almost all algorithm combinations on all problems.  With the exception of 
some combinations using \emph{Single}, all algorithm combinations used over
1,000 nodes.  Half of all combinations without \emph{Single} used 10,000 nodes.
As this was the maximum allowed by our tuning, these combinations may work better with even larger
genomes.

>>>>

In this work we set out to develop a deeper understanding of what makes Cartesian
Genetic Programming (CGP) an effective evolutionary optimizer.
Previous studies of CGP's evolutionary mechanisms have
focused on bloat~\cite{miller:2001:bloat},
neutrality~\cite{vassilev:2000:neutrality}, and structural bias~\cite{payne:2009:bias}.
We shall focus on the interaction between mutation and genome ordering.
%NEW
While these features of CGP have trivial definitions, we will argue, using
experimentation and analysis of alternatives, that they have a great deal
of impact on search success.
%/NEW

@@

As both output locations and information flow in the DAG are evolvable, often
only a tiny fraction of the genome participates in creating the
output values.  These nodes are referred to as ``active,'' with the nodes not
being used to create output values referred to as ``inactive.''  Inactive nodes
allow CGP individuals to drift genetically, as they can be mutated without effecting
the fitness of the individual.  This genetic drift can then be incorporated, as
mutation can change the DAG structure causing previously inactive nodes to become
active.  Previous work suggests CGP is most efficient when up to 95\% of the genome
is inactive~\cite{miller:2006:redundancy}, while more recent work suggests this may
be a result of hidden parsimony pressure in CGP~\cite{goldman:2013:ordering}.

CGP uses very simple evolutionary mechanisms.  The most common evolutionary strategy
is $\mu + \lambda$ where $\mu \leftarrow 1$ and $\lambda \leftarrow 4$.
This means during each generation a single parent produces four offspring using
mutation.  The best offspring then competes with the parent,\footnote{Ties are broken
randomly.} with the offspring
replacing the parent if it is no less fit.  This replacement strategy encourages
neutral drift.
%NEW
In CGP mutation is done in a myriad of nearly synonymous ways.  Here we mutate
%/NEW
each gene of each node at a set probability, where mutation
involves changing a gene randomly to some different valid value.  For example,
if a function gene is chosen for mutation, its new value is randomly chosen from
all possible functions, excluding the gene's current value.
%NEW
This version of mutation was chosen to allow precise prediction of changes
in gene value necessarily in later sections.
%/NEW
Combined,
this form of CGP can be viewed as a stochastic hill climber with neutrality.
For a more in depth description of CGP, see~\cite{miller:2011:chapter2}.

@@

The number of wasted evaluations is
directly related to the mutation probability and the number of active genes.  With
sufficiently high probabilities and enough active genes, the chance of
creating an offspring which is actively identical to its parent is very low.
Yet such a configuration may make it difficult for CGP to optimize, as mutation
is likely to change many active genes in each application.  As was discussed
in~\cite{goldman:2013:cgpwaste}, this can make normal CGP very sensitive to the
mutation probability.

\subsection{Skip}
The \emph{Skip} method for avoiding duplicate evaluations involves the least
amount of modification from CGP's normal behavior.  After an offspring is produced
and its set of active nodes is determined but before it is evaluated, it is compared
with its parent.  Each gene in each active node is compared for equivalence with
the corresponding gene in the offspring's parent.  If all active genes are found to be equal
then the offspring is not evaluated and is instead given the same fitness
as its parent.
Note that since inactive genes are not compared for equivalence it is still possible
for offspring to genetically differ from their parents.

Because \emph{Skip} does not modify any evolutionary mechanisms found in traditional CGP,
it will produce identical individuals, with a potential reduction in evaluations.
If the mutation probability is high enough relative
to the number of active genes, \emph{Skip} will require the same number of evaluations as traditional CGP, since
the probability of an offspring being actively identical to its parent is effectively zero.
When the mutation probability is low enough relative to the number of active genes, some number of
offspring will be actively identical to their parents, resulting in a reduction in evaluations
but no change in the evolutionary trajectory.

%NEW
This technique incurs a minor increase in runtime, as each active gene in the
offspring must be compared against the parent's gene at the same locus.
This operation is at most linear in genome size.  Yet
on problems where evaluation is even somewhat expensive, the evaluations prevented
by performing this check will result in a net reduction in runtime.
%/NEW

We would expect both intuitively and from our initial
experimentation~\cite{goldman:2013:cgpwaste} that \emph{Skip} will be less sensitive
to the mutation probability than normal CGP.  We would also expect \emph{Skip} to use a lower
mutation probability than normal CGP when both are optimized to reduce evaluations.
The former comes from the fact that in normal CGP any mutation probability that has a significant
chance of creating offspring actively identical to their parents wastes
evaluations.  In \emph{Skip} there are only two penalties for reducing the
mutation probability.  From an evolutionary standpoint, if search becomes trapped in a local optima, low mutation
probabilities may have increased difficulty escaping to find the global optimum.  From a performance standpoint,
exceptionally low mutation probabilities may spend a prohibitive amount of time
attempting to produce an evaluable offspring.
Note that these penalties also exist in normal CGP.

@@

Previous experimentation has shown that \emph{Accumulate} acts very similarly to
\emph{Skip}~\cite{goldman:2013:cgpwaste}, with the exception that \emph{Accumulate}
favors lower mutation probabilities when tuned to reduce evaluations to success. While at first these algorithms may appear
quite different, further consideration shows how they are similar.  In \emph{Skip}
any actively identical offspring that is produced is likely to be selected as we expect mutations to
active genes to more often reduce fitness than improve it.  In the next generation
this actively identical offspring is then mutated again, with its lineage likely
continuing until it finally does mutate one or more active genes.  At this point, if the
mutant is better it replaces the parent, otherwise the parent, which has been accumulating
mutations to inactive genes, is kept.  In this way \emph{Skip} mirrors \emph{Accumulate},
except \emph{Skip} acts over multiple generations instead of compressed into a single generation.

%NEW
Similar to \emph{Skip}, \emph{Accumulate} incurs a minor computation cost in
order to prevent wasting evaluations.  Generating each $F_i$ requires at most
$O(AN)$, where $A$ is the arity of the nodes and $N$ is the number of nodes.
While \emph{Accumulate} may produce many $F_i$ in the process of creating
a single offspring, conceptually \emph{Skip} using the same mutation rate
will create a similar number of total individuals when performing the same
number of evaluations.  As such, we suggest \emph{Accumulate}
will take no more than a constant factor more runtime to produce the same number
of evaluable individuals as \emph{Skip}.  As before, this increase is likely
eclipsed when solving problems with sufficiently time consuming evaluation functions.
%/NEW

@@

This modification gives \emph{Single} three properties distinct from the other
forms of duplicate avoidance.  First, \emph{Single} forces offspring to have
an active gene which differs from their parent.  This limits drift as the
mutated active gene must either be to an intron or represent another way to
code a solution of equal quality.  Second, as a benefit of forced changes,
\emph{Single} avoids the overhead of repeatedly generating individuals without
creating an evaluable offspring.  Third, \emph{Single} does not require a mutation
parameter, effectively setting the mutation probability to $\frac{1}{a}$ for active
genes and $\frac{1}{a+1}$ for inactive genes, where $a$ is the number of active genes.

@@

%NEW
Conceptually this method could be extended to mutate some number of active genes
before terminating.  Doing so would reintroduce a parameter to govern how many
active genes to mutate.  Our choice of a single mutation allows for the minimal
amount of change to ensure the individual should be evaluated, while mutations
made to inactive genes still give CGP sufficient power in each mutation to escape
local optimums.
%/NEW

\section{Genome Ordering}
\label{sec:ordering}
As was discussed in Section~\ref{sec:cgp}, CGP uses node ordering in the genome
to prevent cycles, by ensuring nodes only receive input from sources that
precede them in the genome.  This restriction does not limit CGP's ability
to represent DAGs, as all DAGs can be serialized to fit this requirement.
However, this representation is likely having an impact on CGP's ability to evolve specific
DAGs~\cite{goldman:2013:ordering}.

@@

To rectify these issues and examine CGP without ordering bias, \cite{goldman:2013:ordering}
created \emph{Reorder} and \emph{DAG} as methods for genome ordering.
%NEW
For clarity, we shall refer to CGP's historical ordering method of enforcing
only forward connections as \emph{Normal}.
%/NEW

@@

The first step in performing \emph{Reorder} on a genome is to create
data structures to store direct connection relationships.  These structures are used
to determine in constant time all connections incident on a node, and 
can be built in order $O(AN)$ time, where $A$ is the arity of the nodes
and $N$ is the number of nodes.

With this preprocessing complete, the \Call{RandomSerialization}{} algorithm given in Figure~\ref{fig:reorder}
is used to assign each node a new location in the modified genome.
It starts by constructing the \emph{addable} set, which contains all nodes who's direct
dependencies have already been added.  Initially this contains only the input locations.
Once a node is assigned a location, all of the nodes that depend on its output
have that dependency marked as satisfied.  Once all of a node's dependencies
are satisfied, it can be put into \emph{addable}.  In this way, nodes are
randomly removed from \emph{addable}, assigned the next possible location
in the genome, with new nodes added to \emph{addable} as they becomes semantically
viable.  Iteration ends when all nodes have been assigned new locations.
\Call{RandomSerialization}{} requires $O(AN)$, as there can be only $N$ iterations,
with each iteration updating a distinct subset of the $AN$ dependencies.

@@

The price \emph{Reorder} pays for this potential improvement is a larger mutational search
space.  Though a shuffled individual is phenotypically unchanged,
its mutational adjacency can be significantly different.  As such the variety of offspring
an individual can create is increased when using \emph{Reorder}.
%/NEW
Therefore creating a specific offspring that improves upon the parent is
less likely, in that the number of alternatives is increased.  Yet many
of these additional alternatives may also be improvements upon the parent.  This may
increase the likelihood of producing an offspring that improves
upon the parent, allowing \emph{Reorder} to overcome this drawback.
%/NEW

@@

As in \emph{Normal} and \emph{Reorder}, when a connection gene is chosen for mutation, its value is chosen randomly.
However, the options to choose from are only limited to those connections which do not
create cycles.
To achieve this, we incrementally determine
%NEW No change was made to mutating node as that is in fact what we wanted
%/NEW
which nodes transitively depend on the mutating node and which do not, with
information stored in a $known$ map data structure.  This data structure is initialized with
the information that the node is dependent on itself and that all input locations
are not dependent on the mutating node.  Nodes are then
tested in a random order for dependence using the \Call{IsDependent}{} algorithm given in Figure~\ref{fig:dag}.
The first node returned at the top level that is not dependent on the mutating node is then
used as the new connection gene value.  As a result, mutating a connection gene
cannot introduce a cycle.

@@

\Call{IsDependent}{} is a non-repeated recursive depth
first search of the individual's DAG.  Search terminates as soon as a dependent
node is found, and information from previous searches is memoized to prevent
repeated search.  Each recursion level results in a node being added to $known$.
As such initial calls will likely result in lots of recursion, but subsequent
calls will find answers more quickly.  In the worst case, this algorithm may need
to examine all nodes in the genome to determine if $working$ is dependent on the
mutating node, giving \Call{IsDependent}{} a complexity of $O(AN)$.  In the worst case, this algorithm
may be called on all possible nodes in the genome to check their dependence.
Yet cumulatively those calls can only take $O(AN)$ time due to the non-repeating
nature of the search.  As a result, the expected worst case run time for \emph{DAG}
mutation is $O(mA^2N^2)$, where $m$ is the mutation probability.  We expect $mAN$
connection genes to be mutated, each requiring a worst case of $N$ dependency checks
each taking $A$ time.
This is in contrast to \emph{Normal} and \emph{Reorder} which require $O(mAN)$ to perform
mutation, as individual mutations can be performed in constant time.

@@

\begin{table}
	\centering
	\caption{Short descriptions of each of the algorithm variants.}
	\begin{tabular}{|c|p{6cm}|}
	\hline
  \multicolumn{2}{|c|}{\textbf{Duplication}} \\ \hline
	\textbf{\emph{Skip}} & Whenever an actively identical offspring is detected, assign it its parent's fitness. \\ & \\
	\textbf{\emph{Accumulate}} & Repeatedly perform full genome mutation until \textbf{one or more} active genes change. \\ & \\
	\textbf{\emph{Single}} & Select and mutate individual genes until \textbf{exactly one} active gene changes. \\
	\hline \multicolumn{2}{|c|}{\textbf{Ordering}} \\ \hline
	\textbf{\emph{Normal}} & Nodes must take input from other nodes or inputs that precede them in the genome. \\ & \\
	\textbf{\emph{Reorder}} & Each generation the parent's node ordering is shuffled without effecting node behavior. \\ & \\
	\textbf{\emph{DAG}} & Connection genes can mutate to make any connection as long as no cycle is formed. \\
	  \hline
	\end{tabular}
	\label{tab:glossary}
\end{table}

@@

We focus on two parameter values in our configuration: mutation probability and genome size.
While there are other potential configurable parameters (population size, offspring size, etc),
we feel the CGP literature has converged on settings for those parameters (1, 4).
Furthermore, as duplication detection explicitly deals with the relationship between
mutation probability and number of active genes, and the number of active genes depends
on the genome size and ordering method, mutation probability and genome size seem to be
the most likely to impact results.

To set parameter values, we started by defining a grid of parameter values:
50, 100, 200, 500, 1000, 2000, 5000, 10000 for genome size and 0.05, 0.02, 0.01,
0.005, 0.002, 0.001, 0.0005, 0.0002 for mutation probability.  Note that here a genome
size of 50 means there are 50 nodes, each composed of multiple genes, plus the
required genes to specify output locations.  Also note that a mutation probability of 0.05
is used here to mean each gene has a 5\% probability of being mutated in a single
application of mutation.  These ranges were chosen to completely cover the
range of previously used parameter settings.
The grid is formed by trying all possible mutation probabilities
with all possible genome sizes, resulting in 64 potential configurations for each
of the algorithm combinations.  Finally note that as \emph{Single} does not
use a mutation probability, only the eight configurations with different genome sizes were
used when \emph{Single} was employed.

To choose which parameter configuration to use with each algorithm combination
for each problem, we used an iterative process of performing runs, comparing medians, removing configurations,
and repeating.  For each algorithm combination and problem, we first ran all
potential parameter configurations 5 times (320 runs if not using \emph{Single}, 40 otherwise).
The parameter configurations are then sorted based on their median.  The best half are then
each run 4 more times, with the best half of that group chosen.  In this way, 32 configurations
are run 5 times, 16 are run 9 times, 8 are run 13 times, 4 are run 17 times, 2 are run 21 times,
and 2 are run 25 times.  For algorithm combinations using \emph{Single}, all 8 parameter configurations were run 17 times
before any were removed.  This iterative reduction allows us to focus tuning on
those parameter configurations most likely to be effective, with more information
gathered before choosing between high quality configurations.
%NEW
We chose to run each initially 5 times to reduce the chance of
eliminating a configuration due to outliers.  We incrementally added 4
to provide enough information to significantly update estimates,
while keeping the number of runs odd, allowing the median to be a
representative run and not an interpolation.  As a final consideration
the number of experiments were set to make the total number of runs feasible.
%/NEW
In total this means
15,696 complete runs are required to set the parameter configuration for all
algorithm configurations for all problems.

@@

\begin{table*}
	\centering
	\caption{Results for algorithm configuration on each problem.  Highlighted row
	is the control configuration \emph{Skip-Normal}.  MES=Median Evaluations to Success; MAD=Median
	Absolute Deviation; Active=Median number of active nodes in evolved solution;
	Used=Median number of nodes active in evolved solution after applying SIMPLIFY (See Section~\ref{sec:behavior});
	p-value=Result from Mann-Whitney~U comparison with control.
	}
	\begin{tabular}{|c|c|c|c|c|c|c|c|c|}
	  \hline
	  \multicolumn{9}{|c|}{\textbf{3-bit Parity}, Kruskal-Wallis=$5.99\cdot 10^{-14}$} \\ \hline
\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\emph{Accumulate} & \emph{Normal} & 10,000 & 0.02 & 682 & 348 & 122 & 20 & 0.5602 \\ \hline
\emph{Accumulate} & \emph{Reorder}&  5,000 & 0.01 & 796 & 434 & 331 & 23 & 0.7174 \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}       & 10,000 & 0.02 & 823 & 363 & 119 & 20 & NA \\ \hline
\emph{Skip} & \emph{Reorder}      &  2,000 & 0.01 & 870 & 362 & 198 & 20 & 0.3125 \\ \hline
\emph{Accumulate} & \emph{Dag}    &  5,000 & 0.01 & 964 & 499 & 1,174 & 22 & 0.3310 \\ \hline
\emph{Skip} & \emph{Dag}          & 10,000 & 0.02 & 1,042 & 620 & 2,083 & 24 & 0.0248 \\ \hline
\emph{Single} & \emph{Normal}     & 500 & NA & 1,214 & 497 & 39 & 17 & 0.0046 \\ \hline
\emph{Single} & \emph{Reorder}    & 200 & NA & 1,506 & 823 & 57 & 18 & 0 \\ \hline
\emph{Single} & \emph{Dag}        & 100 & NA & 2,569 & 1,313 & 43 & 15 & 0 \\ \hline\hline

@@

\textbf{Duplication} & \textbf{Ordering} & \textbf{Genome Size} & \textbf{Mutation Probability} & \textbf{MES} & \textbf{MAD} & \textbf{Active} & \textbf{Used} & \textbf{p-value} \\ \hline
\rowcolor{Gray}
\emph{Skip} & \emph{Normal}        & 10,000 & 0.005 & 16,226 & 7,455  &   283 & 49 & NA \\ \hline
\emph{Skip} & \emph{Reorder}       & 10,000 & 0.005 & 17,883 & 8,378  & 2,070 & 62 & 0.6918 \\ \hline
\emph{Accumulate} & \emph{Reorder} & 10,000 & 0.002 & 18,132 & 8,926  & 2,027 & 64 & 0.8822 \\ \hline
\emph{Accumulate} & \emph{Normal}  & 10,000 & 0.005 & 21,428 & 6,379  &   279 & 49 & 0.2525 \\ \hline
\emph{Accumulate} & \emph{Dag}     & 10,000 & 0.002 & 24,855 & 12,512 & 4,059 & 69 & 0.0736 \\ \hline
\emph{Skip} & \emph{Dag}           & 10,000 & 0.002 & 25,678 & 12,034 & 4,178 & 62 & 0.0211 \\ \hline
\emph{Single} & \emph{Reorder}     & 100    & NA    & 26,558 & 7,858  &    52 & 32 & 0.0076 \\ \hline
\emph{Single} & \emph{Normal}      & 2,000  & NA    & 31,057 & 14,475 &   141 & 43 & 0.0085 \\ \hline
\emph{Single} & \emph{Dag}         & 100    & NA    & 34,208 & 10,069 &    59 & 34 & 0 \\ \hline\hline

@@

As discussed in Section~\ref{sec:parameter}, the genome size and mutation probability
for each algorithm combination was extensively optimized.  In line with previous
research~\cite{miller:2006:redundancy}, this led to massive genome sizes for
almost all algorithm combinations on all problems.  With the exception of 
some combinations using \emph{Single}, all algorithm combinations used over
1,000 nodes.  Half of all combinations without \emph{Single} used 10,000 nodes.
As this was the maximum allowed by our tuning, these combinations may work better with even larger
genomes.
